---
title: "Homework3.Rmd"
output: html_document
---

Download this file with the command:

```{r, eval=FALSE}
download.file(url = "https://raw.githubusercontent.com/wrightaprilm/Systematics2020/master/Homework/Homework3.Rmd", destfile = "Where you want to save it on your computer")
```

When you are done answering your questions in this document, email it to me.


1. 5 pts. Try remaking the distributions from class, but with a lognormal function (use `rgamma`). In particular, have a look at the scale parameter and the rate parameter. What do these two parameters do? What does a high scale but low rate imply? A low scale bu high rate? Include your code below so I can reproduce the figures.

```{r}
# Code here

```

```
# Descriptive answer here

```


2. 2 points. Have a look at [this](osf.io/jupva) paper. In the section "Assumptions about the evolutionary process," I make a few arguments about how Bayesian and parsimony methods are the same, and how they are different. Do you buy what I'm selling in this section? Why or why not?

```
# Answer here
```


3. 2 points. Something that a lot of researchers struggle with is how to set priors. Based on what you've read so far, how big a problem do you think that is, and do you think researchers can do anything to mitigate damaging effects of not having a strong intuition for the prior?

```

# Answer here

```


4. 1 pts. Continuous trait data aren't well used in phylogenetics right now. Do you think that should change?

```

# Answer here

```
