---
title: "LikelihoodCalc"
author: "April Wright"
date: "2/22/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Let us create a Rate Matrix 


```{r}
row1 <- c(0.0, 1.0, 1.33, 1.0)
row2 <- c(1.0, 0.0, 1.0, 1.33)
row3 <- c(1.33, 1.0, 0.0, 1.0)
row4 <- c(1.0, 1.33, 1.0, 0.0)
rate_mat <- matrix(rbind(c(row1, row2, row3, row4)), nrow = 4)

colnames(rate_mat) <- c("A","C", "G", "T")
rownames(rate_mat) <- c("A","C","G", "T")
rate_mat

```

These are the rates of change. You may see these called exchangeabilities. Now, let's create a vector of base compositions.

```{r}
Pi = c(.4, .1, .2, .3)
diagonal_pi <- diag(Pi)
```

Now we make what is called the Q matrix:


```{r}
Q_unscaled <- rate_mat * Pi
Q_unscaled
```

Next, we want to scale the Q matrix such that the rows sum to zero. 


```{r}
rowsums <- rowSums(Q_unscaled)
rowsums
for (i in 1:4){
  Q_unscaled[i, i] <- -rowsums[1]
}

Q_unscaled
```

Next, we compute a diagonal sum for the matrix. This is used to normalize your Q matrix.

```{r}
offdiags <- function(mat = matrix){
diag_sum <- sum(diag(mat))
off_diag_sum <- sum(i) - diag_sum
return(off_diag_sum)
}


```


First we transform by the matrix diagonal of pi.
```{r}
b <- Q_unscaled * diagonal_pi
b
```


And this gives us a scaling constant.
```{r}

scaling <- offdiags(b)

scaledQ <- Q_unscaled/scaling


```


Now we use this to calculate a likelihood of observing a particular branch length per site:

Pr(Branch Length per site) = scaledQ * BranchLength. For example, if we have our Q matrix, and the branch length is .1, the below are the probabilities of seeing a change between those characters on that branch:

```{r}
longish_branch <- scaledQ*.05

```


Now lets imagine we have two taxa, and we have an alignment like so:

Taxon1 = CCAT
Taxon2 = CCGT

The probability will look like this:


```
Likelihood(Data | Model) = Probability(C) * Pr(C --> C) * Pr(C) * Pr(C-->C) * Pr(A) * Pr(A --> G) * Pr(T)* Pr(T-->T) 
```

In code, this will be

```{r}
Pi[2]*longish_branch[2,2] * Pi[2]*longish_branch[2,2] * Pi[1] * scaledQ[1,3]*Pi[4]*longish_branch[4,4]

```


# Priors

Priors reflect a prior belief about a variable in the analysis. In this case, for example, we may believe a branch length to be drawn from an exponential distribution. Take a look here at the exponential: 

```{r}

branch_length_prior <- rexp(1000000, 10)
hist(branch_length_prior)

```

Let's try the above calculation with a prior included.

```{r}

branch_length_prior <- rexp(1, 10)

prior_branch <- scaledQ*branch_length_prior
Pi[2]*prior_branch[2,2] * Pi[2]*prior_branch[2,2] * Pi[1] * scaledQ[1,3]*Pi[4]*prior_branch[4,4]

```

